jax-cosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-01-24 paper.tex
add a figure to show bias correlation. Start Appendix for DES-Y1 paameters...

If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$). So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.
%
\subsubsection{DES-Y1 Example}
%
Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI. The Adam optimizer has been used with a learning rate set to $10^{-3}$.  Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler. The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a V100 GPU. Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (vectorized) with 1000 samples each. The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively. Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running. The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}. The remarkable outcome is that within few hours on a single GPU, one can figure out progressively if the contours have a meaningful appearance before launching multi-GPU multi-chains NUTS sampling to get high statistics Markov Chain with high effective sample size (ESS). Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}....
\begin{figure}
\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200.pdf}
\caption{Constraints on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in..., and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
\label{fig_svi_nuts_1}
\end{figure}

@@

\begin{table}[htb]
\caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}. (1) are computed from MCMC obtained by the the NUTS sampler described..., while (2) are the results of SVI followed by NUTS sampler after a Neural Transform  with 10 chains of 1000 sampling each after 200 warmup samples.}
\label{tab-ESS-NUTS_SVI-1}
 \centering
\begin{tabular}{ccccccccccc}
\hline
         & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$     & $n_s$ & $A$ \\
\hline
(1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ \\
(2) & $38.7$ &  $24.8$     & $16.0$     & $18.6$ & $43.3$ & $39.5$ & $21.7$ \\
\end{tabular}
\end{table}

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

>>>>

If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$). So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.
%
\subsubsection{DES-Y1 Example}
%
Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI. The Adam optimizer has been used with a learning rate set to $10^{-3}$.  Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler. The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a V100 GPU. Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (vectorized) with 1000 samples each. The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively. Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running. The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}. The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.

The remarkable outcome is that within few hours on a single GPU, one can figure out progressively if the contours have a meaningful appearance before launching multi-GPU multi-chains NUTS sampling to get high statistics Markov Chain with high effective sample size (ESS). Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}....
\begin{figure}
\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200.pdf}
\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 22 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in..., and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
\label{fig_svi_nuts_1}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200_bis.pdf}
\caption{In the same conditions of Fig.~\ref{fig_svi_nuts_1} but for the five lens galaxy bias which are highly correlated.}
\label{fig_svi_nuts_4}
\end{figure}

\begin{table}[htb]
\caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}. (1) are computed from MCMC obtained by the the NUTS sampler described..., while (2) are the results of SVI followed by NUTS sampler after a Neural Transform  with 10 chains of 1000 sampling each after 200 warmup samples.}
\label{tab-ESS-NUTS_SVI-1}
 \centering
\begin{tabular}{ccccccccccc}
\hline
         & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$ & $n_s$ & $A$ & $\eta$\\
\hline
(1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\
(2) & $38.7$ &  $24.8$     & $16.0$     & $18.6$ & $43.3$ & $39.5$ & $21.7$ & $16.6$\\
\hline
\end{tabular}
\end{table}

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
\appendix

\section{Description of the DES-Y1}
\label{app-DESY1}
\begin{table}[htb]
\caption{Priors of the 22 variables of the DES-Y1 of the 3x2pt likelihood (number counts and shear).}
\label{tab-DESY1}
 \centering
\begin{tabular}{ccccccccccc}
\hline
 parameter &  prior \\
 \hline
  \multicolumn{2}{c}{\textbf{Cosmology}} \\
$\Omega_b$ & $\mathcal{U}[0.10, 0.9]$ \\
$\Omega_c$ & $\mathcal{U}[0.40, 1.0]$ \\
$\sigma_8$ & $\mathcal{U}[0.03, 0.07]$ \\
$w_0$ & $\mathcal{U}[-2.00, -0.33]$ \\
$h$ & $\mathcal{U}[0.55, 0.91]$ \\
$n_s$ & $\mathcal{U}[0.87, 1.07]$ \\
\multicolumn{2}{c}{\textbf{Intrinsic Alignment}} \\
$A$ & $\mathcal{U}[-5,5]$ \\
$\eta$ &$\mathcal{U}[-5,5]$ \\
\multicolumn{2}{c}{\textbf{Lens Galaxy Bias}} \\
$(b_i)_{i:1,\dots,5}$ & $\mathcal{U}[0.8,3.0]$ \\
\multicolumn{2}{c}{\textbf{Shear Calibration Systematics}} \\
$(m_i)_{i:1,\dots,4}$ & $\mathcal{N}[0.012,0.023]$ \\
\multicolumn{2}{c}{\textbf{Source photo-$z$ shift}} \\
$dz_1$ & $\mathcal{N}[0.001,0.016]$ \\
$dz_2$ & $\mathcal{N}[-0.019,0.013]$ \\
$dz_3$ & $\mathcal{N}[0.009,0.011]$ \\
$dz_4$ & $\mathcal{N}[-0.018,0.022]$ \\
\hline
\end{tabular}
\end{table}
